Tesseroids: command-line tools for gravity field forward modeling in spherical coordinates

### 2014-04-01 manuscript.tex
Writing methods until adaptive discretization part
Added Hildebrand reference for the GLQ and edited equations.

\address{
    \footnotemark[1]
    Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil,
    e-mail: \\
    \footnotemark[2]
    Observat\'orio Nacional, Rio de Janeiro, Brazil,
    e-mail:
}

@@

Formula by \citet{Grombein2013}.

First appears in \citet{Asgharzadeh2007}.

Studied by \citet{Wild-Pfeiffer2008}.

Numerical accuracy by \citet{Ku1977}.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$.
    }
    \label{fig:tesseroid}
\end{figure}

This is a reference to Figure~\ref{fig:tesseroid}.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid-split.png}
    \caption{
        Adaptive discretization for a computation point P
        of the tesseroid shown in (a) using distance/size ratios of
        (b) 0.5, (c) 1, and (d) 2.
        $d$ is the distance between the tesseroid and P.
    }
    \label{fig:split}
\end{figure}

This is a reference to Figure~\ref{fig:split}.

>>>>

\address{
    \footnotemark[1]
    Universidade do Estado do Rio de Janeiro, Rio de Janeiro, Brazil,
    e-mail: leouieda@gmail.com;\\
    \footnotemark[2]
    Observat\'orio Nacional, Rio de Janeiro, Brazil;
}

@@

The gravitational potential,
gravitational acceleration,
and Marussi (gravity gradient) tensor
caused by a homogeneous tesseroid
with density $\rho$
are \citep{Grombein2013}

\begin{equation}
    V(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{1}{\ell} \kappa  dr' d\phi' d\lambda',
    \label{eq:tesspot}
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        \frac{\Delta_{\alpha}}{\ell^3} \kappa dr' d\phi' d\lambda'
        \ \ \ \alpha \in \{x,y,z\},
    \label{eq:tessgrav}
\end{equation}

\noindent
and

\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) = G \rho
        \int\limits_{\lambda_1}^{\lambda_2}
        \int\limits_{\phi_1}^{\phi_2}
        \int\limits_{r_1}^{r_2}
        I_{\alpha\beta}
        dr' d\phi' d\lambda',
    \label{eq:tesstensor}
\end{equation}

\noindent
in which
$(\phi, \lambda, r)$ are
the latitude, longitude, and radius
coordinates of computation point P (Figure~\ref{fig:tesseroid}),
$\alpha,\beta \in \{x,y,z\}$
(the local coordinate system of point P),
and

\begin{eqnarray}
    I_{\alpha\beta} &=&
        \left(
            \frac{3\Delta_{\alpha} \Delta_{\beta}}{\ell^5} -
            \frac{\delta_{\alpha\beta}}{\ell^3}
        \right) \kappa, \\
    \Delta_x &=& r' K_{\phi} , \\
    \Delta_y &=& r' \cos \phi' \sin(\lambda' - \lambda) , \\
    \Delta_z &=& r' \cos \psi - r, \\
    \ell &=& \sqrt{r'^2 + r^2 - 2 r' r \cos \psi} , \\
    \cos\psi &=& \sin\phi\sin\phi' + \cos\phi\cos\phi'
                 \cos(\lambda' - \lambda) , \\
    K_{\phi} &=& \cos\phi\sin\phi' - \sin\phi\cos\phi'
                 \cos(\lambda' - \lambda), \\
    \kappa &=& {r'}^2 \cos \phi'.
\end{eqnarray}

\noindent
$\delta_{\alpha\beta}$ is the Kronecker delta,
i.e. $\delta_{\alpha\beta}=1$ if $\alpha=\beta$
and $\delta_{\alpha\beta}=0$ otherwise.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid}
    \caption{
        View of a tesseroid,
        the integration point $Q$,
        a geocentric coordinate system $(X, Y, Z)$,
        the computation $P$ and it's local coordinate system $(x, y, z)$.
        $r$, $\phi$, $\lambda$ are
        the radius, latitude, and longitude, respectively, of point $P$,
        and $\ell$ is the Cartesian distance between $P$ and $Q$.
    }
    \label{fig:tesseroid}
\end{figure}

The above integrals
have to be solved numerically
\citep{Wild-Pfeiffer2008}
and different approaches have been proposed.
\citet{Heck2007}
used a Taylor series expansion,
while \citet{Asgharzadeh2007}
used the Gauss-Legendre Quadrature (GLQ).
\citet{Wild-Pfeiffer2008} investigated
the use of both methods,
along with alternative mass elements,
and found the GLQ to be the most efficient method.
Therefore, we have opted to use
the 3D GLQ in our computations.

The Gauss-Legendre Quadrature
approximates the integral by
a weighted sum
\citep{Hildebrand1987},
e.g.

\begin{equation}
    \int\limits_a^b f(x) dx \approx
    \frac{b-a}{2}\sum\limits_{i=1}^N W_i f(x_i),
\end{equation}

\noindent
where the discretization points (nodes) $x_i$
are the roots of the $N$th order Legendre polynomial $P_N$
and the weights $W_i$ are \citep{Hildebrand1987}

\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P'_N(x_i))^2}.
\end{equation}

The triple integrals in equations
\ref{eq:tesspot},
\ref{eq:tessgrav},
and
\ref{eq:tesstensor}
become

\begin{equation}
    V(r,\phi,\lambda) \approx
        A
        \sum\limits_{k=1}^{N^{\lambda}}
        \sum\limits_{j=1}^{N^{\phi}}
        \sum\limits_{i=1}^{N^r}
        W^r_i W^{\phi}_j W^{\lambda}_k
        \frac{1}{\ell} \kappa,
\end{equation}
\begin{equation}
    g_{\alpha}(r,\phi,\lambda) \approx
        A
        \sum\limits_{k=1}^{N^{\lambda}}
        \sum\limits_{j=1}^{N^{\phi}}
        \sum\limits_{i=1}^{N^r}
        W^r_i W^{\phi}_j W^{\lambda}_k
        \frac{\Delta_{\alpha}}{\ell^3} \kappa,
\end{equation}

\noindent
and

\begin{equation}
    g_{\alpha\beta}(r,\phi,\lambda) \approx
        A
        \sum\limits_{k=1}^{N^{\lambda}}
        \sum\limits_{j=1}^{N^{\phi}}
        \sum\limits_{i=1}^{N^r}
        W^r_i W^{\phi}_j W^{\lambda}_k
        I_{\alpha\beta},
\end{equation}

\noindent
where

\begin{equation}
    A = G \rho
    \frac{(\lambda_2 - \lambda_1)(\phi_2 - \phi_1)(r_2 - r_1)}{8},
\end{equation}

\noindent
$W_i^r$, $W_j^{\phi}$, and $W_k^{\lambda}$
are the weights
and $N^r$, $N^{\phi}$, and $N^{\lambda}$
are the number of nodes
for the radial, latitudinal, and longitudinal dimensions, respectively.

The accuracy of the numerical integration
using the GLQ
was investigated by \citet{Ku1977}
for the case of the right-rectangular prism.
\citet{Ku1977} found that the accuracy
depends on the ratio between
the average distance between GLQ nodes
and the distance to the observation point P.
As a rule-of-thumb,
\citet{Ku1977} suggests that
the distance between nodes
should not be greater than
the distance to point P.
\citet{Li2011} devised a scheme that
recursively subdivides the tesseroids
into smaller ones,
effectively decreasing the distance between nodes.
This scheme guarantees that
the rule-of-thumb is not violated.
Hence,
optimal accuracy of the numerical integration
is achieved automatically.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figs/tesseroid-split}
    \caption{
        Adaptive discretization
        of the tesseroid shown in (a)
        for a computation point P
        using distance/size ratios of
        (b) 0.5, (c) 1, and (d) 2.
        $d$ is the distance between the tesseroid and P.
    }
    \label{fig:split}
\end{figure}
